---
title: "Workshop Regression -- Practice"
author: "A.Ponsero"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---


```{r setup, message=FALSE, warning=FALSE}
library(tidymodels)      # For modeling
library(palmerpenguins)  # For our penguin data
library(tidyverse)       # For data manipulation and visualization

# Resolve any function conflicts in favor of tidymodels
tidymodels_prefer()
```


## 7. Practice Exercises

Now it's your turn! These exercises will help you practice what you've learned. Start with the basics and work your way up to the challenges.

## Exercise 1: Build Your Own Simple Model

Build a linear regression model to predict **bill length** from **bill depth**.

**Tasks:**

a) Create a train/test split (75/25) using the clean penguins data. Set seed to 456.

b) Fit a simple linear regression model predicting `bill_length_mm` from `bill_depth_mm`

c) Calculate RMSE and R² on both training and test sets

d) Create a visualization showing the fitted line with the data points

```{r exercise-1, eval=FALSE}
# Your code here
set.seed(456)

# a) Create split


# b) Fit model


# c) Calculate metrics


# d) Visualize
```

**Questions to consider:**
- Is bill depth a good predictor of bill length?
- How does this R² compare to our body mass models?

Hint:
```{r exercise-1-hints, eval=FALSE}
# Hint for (a)
bill_split <- initial_split(penguins_clean, prop = ___, strata = ___)

# Hint for (b)
lm_model %>%
  fit(___ ~ ___, data = ___)

# Hint for (c)
# Use penguin_metrics() we created earlier
```


---

## Exercise 2: Exploring Different Predictors

We used flipper length as our main predictor. Let's see how other single predictors perform!

**Task:** Build separate simple models predicting body mass from:
1. Bill length only
2. Bill depth only
3. Compare their test set RMSE to our flipper length model (which had RMSE ≈ ???g)

```{r exercise-2, eval=FALSE}
# Your code here

# Model 1: Bill length


# Model 2: Bill depth


# Compare RMSEs
```

**Questions:**
- Which single predictor works best?
- Does this match what we saw in the exploratory data analysis?
- Why might bill depth perform worse than expected?

---

## Exercise 3: Adding an Interaction Term

Species and sex both affect body mass, but does their **effect interact**? Maybe the sex difference is larger in some species than others.

**Task:** Create a model with an interaction between species and sex.

a) Build a recipe that includes the interaction term
Hint: Use `step_interact(~ species:sex)` after creating dummy variables

b) Fit the model and examine coefficients

c) Compare test set performance to the model without interactions

```{r exercise-3, eval=FALSE}
# Your code here

# Create recipe with interaction
penguin_recipe_interact <- recipe(body_mass_g ~ flipper_length_mm + 
                                    bill_length_mm + bill_depth_mm + 
                                    species + sex, 
                                  data = penguin_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_interact(~ ___:___) %>%  # Add interaction
  step_zv(all_predictors())

# Fit and evaluate
```

**Questions:**
- Did adding the interaction improve performance?
- What does the interaction coefficient tell you biologically?

Hint:
```{r exercise-3-hints, eval=FALSE}
# Interaction syntax
step_interact(~ starts_with("species"):starts_with("sex"))

# Or more explicitly
step_interact(~ species_Chinstrap:sex_male + species_Gentoo:sex_male)
```

---

## Exercise 4: Investigating Prediction Errors

Let's dig deeper into where our multiple regression model makes mistakes.

**Tasks:**

a) Using the test set results from our multiple regression model, identify the 5 penguins with the **largest absolute errors**

b) For these penguins, create a table showing:
   - Species, sex, actual body mass, predicted body mass, error
   - All predictor variables

c) Can you spot any patterns? Are errors concentrated in certain groups?

d) Create a visualization showing prediction errors by species

```{r exercise-4, eval=FALSE}
# Your code here

# a) Find worst predictions


# b) Detailed table


# c) Look for patterns


# d) Visualize errors by species
# Hint: Use a boxplot of residuals grouped by species
```

**Reflection questions:**
- Are we systematically over- or under-predicting for any species?
- Which species has the most variable prediction errors?
- What might explain these patterns?

---

## Exercise 5: Comparing All Models

Create a comprehensive comparison of all models we've built (or you've built in exercises).

**Task:** Create a publication-ready table and/or visualization comparing:
- Simple model (flipper only)
- Multiple regression (all predictors)
- Any other models from your exercises

Show both training and test metrics for each.

```{r exercise-5, eval=FALSE}
# Your code here

# Collect all model results


# Create comparison table


# Create comparison visualization
```

---

## Exercise 6: Feature Engineering

Can you improve the model by creating new features?

**Ideas to try:**

a) Create a **bill ratio**: `bill_length_mm / bill_depth_mm`

b) Create a **body shape index**: Could combine multiple measurements

c) Create **polynomial terms**: Maybe flipper_length² helps?

**Task:** 
1. Add your engineered features to a recipe
2. Fit the model
3. Evaluate on test set
4. Did your features improve performance?

```{r exercise-6, eval=FALSE}
# Your code here

# Example: Add derived features in recipe
penguin_recipe_engineered <- recipe(body_mass_g ~ ., 
                                    data = penguin_train) %>%
  step_mutate(
    bill_ratio = bill_length_mm / bill_depth_mm,
    # Add your features here
  ) %>%
  # Continue with other preprocessing
```

**Challenge questions:**
- Which engineered features help most?
- Can you achieve R² > 0.90 on the test set?
- Is a more complex model always better?

---

## Exercise 7: Prediction Challenge

**Real-world scenario:** A field researcher measures these penguins but doesn't have a scale to weigh them. Use your best model to predict their body mass!

```{r challenge-penguins}
# Mystery penguins
mystery_penguins <- tibble(
  penguin_id = 1:5,
  species = c("Adelie", "Gentoo", "Gentoo", "Chinstrap", "Adelie"),
  island = c("Torgersen", "Biscoe", "Biscoe", "Dream", "Dream"),
  sex = c("male", "female", "male", "female", "male"),
  flipper_length_mm = c(191, 220, 230, 196, 185),
  bill_length_mm = c(40.5, 49.1, 50.8, 46.3, 37.2),
  bill_depth_mm = c(18.9, 14.8, 15.9, 17.8, 19.5)
)
```

**Tasks:**

a) Use your best model to predict body mass for each penguin

b) Provide a **prediction interval** or estimate of uncertainty for each prediction
   - Hint: Think about your model's RMSE - that's the typical error

c) Which predictions are you most/least confident about, and why?

```{r exercise-8, eval=FALSE}
# Your code here

# Make predictions


# Add uncertainty estimates
```

---

## Exercise 9: Explore on Your Own

**Open-ended exploration!** Choose one or more:

**Option A: Different research question**
- Instead of predicting body mass, can you predict flipper length from other variables?
- How does model performance compare?

**Option B: Species-specific models**
- Build separate models for each species
- Do different predictors matter for different species?
- Compare to the model with species as a predictor

**Option C: Temporal analysis**
- Does body mass change across years (2007-2009)?
- Could you build a model that accounts for year?

**Option D: Investigate Simpson's Paradox**
- Create visualizations showing how bill depth relates to body mass
- Show the relationship overall vs. within species
- Explain why the relationship "reverses"

---

## Reflection Questions 🤔

After completing the exercises, reflect on:

1. **What was the hardest part of building regression models?**
   - Data splitting? Model specification? Interpretation?

2. **When would you use a simple vs. complex model?**
   - What are the tradeoffs?

3. **How would you explain train/test splitting to a colleague?**
   - Can you use a non-technical analogy?

4. **What surprised you most about the penguin data?**
   - Any unexpected relationships?

5. **How might you apply these techniques to your own research data?**
   - What would be your outcome variable?
   - What predictors might you use?

---

## Solutions and Discussion

Solutions will be discussed as a group. Before looking at solutions:
- Try the exercises yourself first!
- Discuss with a neighbor
- Don't worry about getting everything perfect

> 💡 **Remember**: In real research, there's rarely one "right" answer. Different approaches have different tradeoffs, and the best choice depends on your specific research question and goals!

---

## Additional Resources

Want to learn more? Check out:

**Books:**
- [Tidy Modeling with R](https://www.tmwr.org/) by Max Kuhn & Julia Silge
- [R for Data Science](https://r4ds.had.co.nz/) by Hadley Wickham & Garrett Grolemund

**Videos:**
- [Julia Silge's YouTube channel](https://www.youtube.com/c/JuliaSilge) - Excellent tidymodels tutorials

**More about penguins:**
- [Palmer Penguins package website](https://allisonhorst.github.io/palmerpenguins/)
