---
title: "Challenge_answers"
author: "Brooke Wolford"
date: "2025-12-04"
output: html_document
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load packages
library(dplyr)
library(readr)
library(tidyr)
```

# Challenge Answers

Task: Given a simple data frame of patients with ID, sex, age, and cholesterol values in two visits, write code to:
	1.	Filter patients over 45 years and save to a new dataset called sub, how many are there?
	2.  Select ID, Sex, and Cholesterol columns to a new dataset called sub2.
	3.	Using sub2, create a new variable that is 1 if the patient has high cholesterol (over 200), and 0 otherwise. Save this to a datset called high.
	4.	Using sub2, summarize average cholesterol by sex

```{r}

df <- data.frame(
  ID = 1:8,
  Age = c(34, 60, 45, 50, 29, 80, 55, 65),
  SBP = c(120, 140, 135, 150, 110, 140, 120, 125 ),
  cholesterol = c(200,220, 190, 195, 180, 240, 230, 235),
  sex = c("Male","Male","Female","Male","Female","Female","Male","Male")
)

# Filter patients over 45 and save to a new dataset named sub
sub<-df %>% filter(Age>45)

# How many patients are over 45?
nrow(sub)

# In the original dataset, select the ID, sex, and cholesterol column for a new dataset
sub2<-df %>% select(ID,sex,cholesterol)

# Create a new variable (indicator variable) showing whether cholesterol is above 200
sub3<-sub2 %>% mutate(high=ifelse(cholesterol>200,1,0))

# Summarize average cholesterol by sex
sub2 %>% group_by(sex) %>%summarise(mean_chol=mean(cholesterol))


```
